Infant Birth Weight in Brazil

### 2024-09-03 products/manuscript/Manuscript_Final_Lancet.qmd
Revisions for submission 5
clean up of code and text

In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence. After abolition, the state did not create programs to incorporate people formerly enslaved and their descendants into economic, civil, or political life. Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century. Public health studies have shown that health disparities continue to mirror patterns from over a century ago. The goal of this study is to quantify health disparities in infant birth weights by maternal characteristics in the post-abolition period in one of Brazil's first public maternity hospitals.

@@

This cross-sectional analysis used hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital. I used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.

@@

Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. In the field of maternal-infant health, both low birth weight (LBW) and preterm birth (PTB) within and across communities is a focal point of research [@hughes2017; @silva1998; @wilcox2001]. LBW can be a function of PTB, of intra-uterine growth retardation (IUGR) (otherwise known as small-for-gestational-age or SGA) or of both [@hughes2017].

LBW is associated with higher rates of neonatal and infant mortality [@paixao2021; @vilanova2019]. Longitudinal studies also have demonstrated an association between at-term LBW and increased risk for cardiovascular diseases and other chronic conditions over the life course [@barker2004]. But birth weight in general and LBW in particular are often proxies for unmeasurable variables. Thus, these categories should not be understood as a causal mechanism of outcomes such as infant mortality or later-life chronic disease; rather, they provide important population-level information [@wilcox2001].

Contemporary studies in Brazil have shown that non-White mothers have higher odds of giving birth to LBW infants [@nyarko2013; @wehby2015a; @vilanova2019]. Non-White mothers also had infants with lower mean birthweights than their White counterparts [@leal2006]. These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns are at higher risk for LBW than White newborns, with trends worsening over time [@pollock2021]. Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997].

@@

Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere. In the nineteenth century, population-level health was poor for all Brazilians, but particularly so for enslaved individuals, who not only experienced the epidemic diseases such as cholera, plague, and yellow fever that affected the entire population, but also suffered from nutritional deficiencies [@karasch1987; @kiple1989]. EXPAND

@@

By the 1920s, at least in Brazil’s large cities, public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @martins2004; @roth2020]. Dedicated maternity hospitals such as Laranjeiras were central to this hospitalization process. There, obstetricians and gynecologists began implementing surgical advances in labor and delivery [@martins2004; @roth2021]. Underlying these structural and technological changes was an ideology scholars have described as "scientific motherhood," supported by both physicians and elite women, which relied on a technocratic model towards pregnancy and motherhood and, in doing so, essentialized notions of gender and foregrounded physicians’ scientific authority [@freire2009; @martins2004; @otovo2016].

At the same time, longstanding racial hierarchies forged under racial slavery morphed into "scientific" ways to categorize humans based on skin color in post-abolition contexts. In Brazil, with a large Black and mixed-race population, early twentieth-century racist thinkers first embraced a theory of "whitening" in which they believed that racial mixture would eventually lead to a White population and then more preventive eugenic theories in the 1920s and 1930s that focused on improving environmental factors rather than restricting reproduction through sterilization [@stepan1991]. Although some eugenicists called for sterilization of citizens they deemed unfit, they never couched these discussions in explicitly racial terms [@wegner2013]. Whiteness remained a cultural and social goal even if the state did not engage in coercive measures of reproductive control.

Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, was ground zero for the intersection of these forces. Laranjeiras provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020]. Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe. Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021]. Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.

BILL COMMENT: I AM CURIOUS, DO WHITE SOUTH AFRICANS HAVE HIGHER BIRTH WEIGHT, CONTROLLING FOR INCOME? ANY STUDY OF THIS?

This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993]. It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].

@@

This study analyzes infant birth weight as a proxy for general fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras. It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives. BILL COMMENT: PUT THIS PARAGRAPH MUCH EARLIER.

@@

This analysis draws on a unique data set of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926. I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). In the initial months of publication, the reports included more complete information, including labor and delivery time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a]. BILL COMMENT: FOR OBSTETRICIANS WHO APPEAR MORE THAN ONCE, IS THERE A RELATIONSHIP WITH BIRTH WEIGHT?

@@

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only. In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022]. We can question the use of this cutoff in historical studies, as it was not an established category until the mid-twentieth century. A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification. This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity [@wilcox CITE CORRECTLY]. The WHO adopted this cutoff in 1948, and it has remained the standard since [@hughes2017].

@@

*Maternal nationality*. The original hospital records recorded maternal nationality by country, which are available in the supplemental file. For analysis here, I created categories based on individual country (`nationality`) and a dummy variable based on region (`modifiednationality`). I recategorized the latter as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian).

@@

*Infant length*. Infant length was recorded in centimeters in the original records. A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021]. The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009]. But other health problems related to maternal undernutrition during pregnancy depend on gestational period. Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].

@@

Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts, both of color and White [@franken2019; @ball2020]. Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 in two steps. First, I created a new variable `Brazilian` (yes/no) and ran a simple linear analysis with birth weight as the dependent variable and maternal nationality as the independent variable. Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value <0.001). Given that all women of color in the data set were Brazilian, there is collinearity occurring. Thus, second, I dropped all non-Brazilians from the study, and ran the original regressions from Models 1 and 2.

@@

The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers. A difference in means test showed borderline significance (t-statistic = 1.9119, df. 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]). Regardless of a difference between infants born to mothers of different skin colors, all LBW infants in the past, the probability of surviving the neonatal period was lower than today given higher neonatal and infant mortality rates [@costa1998; @roth2020].

Results for all linear models are displayed in Tables 2 and 3. All models demonstrate an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race). When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter than their Euro-descended counterparts (Table 2). In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers. Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent. AT THIS POINT I NEED TO SEE A LITERATURE THAT TELLS ME 100 GRAMS MATTERS FOR SOMETHING, HEALTH-WISE, IS 100 GRAMS MORE THAN 2 STANDARD DEVIATIONS?

UPDATE FOR CORRELATION STUFF

In models 1 and 2 (Table 2), women who were nulliparous (delivering their first child) had infants weighing on average 113 grams less than multiparous women. In models 3 and 4 (Table 3), the average was 114 grams lighter. This is an association found both in other historical studies and today [@costa1998; @souza2020]. Age and nationality were not significant. UPDATE

@@

This study found that women of African descent in early twentieth-century Brazil both had infants with lower birth weights and higher rates of LBW than their White counterparts. When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today. Although the mean birth weight was still within today's normal range, the differences between racialized groups demonstrate differential health outcomes that have persisted until today.

Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight across the Americas and in New Zealand.

Black infants born at Laranjeiras had the lowest mean birth weight (3037 grams) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004]. Black infants born at Laranjeiras were almost 150 grams lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period. They were less than 100 grams lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. White infants born at Laranjeiras had a mean birth weight of 3133 grams, slightly more than 50 grams lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. They were nearly 300 grams lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 grams). As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same historical period [@costa1998; @costa2004; @goldin1989].

What can explain the differences in birth weight both historically and today? The intrauterine environment is a critical determinant of infant birth weight. Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care can all influence birth weight. Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986]. Other factors include maternal age, parity, birth spacing, and gestational age. We can imagine that poorer women in early twentieth-century Rio de Janeiro performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992]. These same women might have lived in crowded housing arrangements with lack of access to running water [@graham1992; @fischer2008; @roth2020]. Generalized poverty might have influenced both maternal nutrition and the ability to access prenatal care.

WHO: birth weight "is strongly conditioned by the health and nutritional status of the mother"

WARD: "In addition, because birth weight is sensitive to environmental factors during the nine months of gestation (especially the final three months) as well as the long-range effects of changes in living conditions, it reflects both short- and long-term trends." (5)

WARD: "By and large, the most privileged social groups bear the heaviest babies while the most disadvantaged have the lightest. These differences, in turn, reflect the fact that fetal growth and weight at birth are influenced by a wide range of biological, social, and economic factors, not all of which are fully understood." (10)

WARD: From the standpoint of neonatal mortality the optimum weight of a newborn child lies between 3500 and 3900 grams. Infants bom outside this range have higher mortality levels than do those within it. The riSks are especially great among those weighing considerably less than the optimum. Estimates suggest that the chance of death during the first week of life increases by 40 percent, and the risk of fetal death by 25 percent, for every 100-gram decrease below the optimum level of weight. (10)

Optimal birth weight is the weight at which birth weight-specific mortality is lowest

@@

All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014; @ward1993]. In this study, a major limitation was the original published clinical notes' exclusion of gestational age. Thus, the birth weight variable captured both smaller babies born at term (low weight at-term infants, or small for gestational age, SGA) and pre-term infants, or those born before 37 weeks' gestation. We are thus unable to know how many of these infants were low birth weight at term [@falcao2020]. This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@cutland2017]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have datasets that include gestational age [@costa1998; @costa2004; @goldin1989]. I assume that infants at risk for PTB were also at-risk for LBW.

@@

HOW INFANTS WERE WEIGHED AND ACCURACY OF SCALE

>>>>

In 1888 Brazil became the last country in the Western Hemisphere to abolish slavery after nearly four centuries of its existence. After abolition, the state did not create programs to incorporate formerly enslaved people and their descendants into economic, civil, or political life. Qualitative historical research has outlined the gendered and racialized health disparities of people of African descent at the turn-of-the-twentieth century. Public health studies have shown that health disparities continue to mirror patterns from over a century ago. The goal of this study is to quantify health disparities in infant birth weights by maternal characteristics in the post-abolition period in one of Brazil's first public maternity hospitals.

@@

This cross-sectional analysis used hospital records on infants delivered between 1922-1926 at the Maternidade Laranjeiras (now Maternidade Escola) in Rio de Janeiro, Brazil, the city's first public maternity hospital. It used linear models to assess differences in infant birth weight by maternal skin color, age, number of pregnancies, and nationality.

@@

Birth weight and its relationship to infant health have been a focus of public health research for over a century [@hughes2017]. In epidemiology, both low birth weight (LBW) and preterm birth (PTB) within and across racialized communities is a focal point of research [@braveman2021a; @david1997; @hughes2017; @silva1998; @wilcox2001]. LBW can be a function of PTB, of fetal growth restriction (FGR, previously known as intra-uterine growth restriction or IUGR, which can include but is not limited to small-for-gestational-age or SGA infants -- those born at less than 10 percentile), or of both [@wilcox2001; @wilcox2024]. Nonetheless, PTB is the primary determinant of both LBW and lower birth weight in general [@maruyama2020].

Multiple social, economic, biological, and genetic factors influence birth weight, which is also partially conditioned by maternal nutrition, health, and health care utilization during pregnancy [@braveman2021a; @david1997; @ward1993]. Recent genetic studies have detailed the specific alleles through which both the maternal and fetal genomes contribute to birth weight and gestational length, while acknowledging the continued importance of non-genetic factors [@juliusdottir2021; @sole-navais2023; @warrington2019]. Given that these studies are based on well-off populations in countries with low neonatal and infant mortality rates, they might overestimate genetic impact on birth weight for historical studies of impoverished populations.

Although not a causal mechanism, LBW is associated with higher rates of neonatal and infant mortality [@paixao2021; @vilanova2019]. Originally, epidemiologists hypothesized an association between at-term LBW and increased risk for cardiovascular diseases and other chronic physical health conditions over the life course [@barker2004]. Yet, more recent genetic research and epidemiological studies employing instrumental variable (IV) analysis contest this. For example, Warrington et al [@warrington2019] argue that genetic effects and not intrauterine programming are associated with later life high blood pressure. Maruyama and Heinesen [@maruyama2020] have found that lower birth weight is associated with both infant mortality and other health issues in the neonatal period, but its association with long-term health issues declines over time.

Historians have shown how birth weight can be an indicator for the often unmeasurable nutritional status of mothers of different social classes and racial classifications in the past [@goldin1989; @costa1998; @costa2004; @ward1993]. Historically, women from more privileged groups had heavier infants than women from less advantaged groups [@stein1984; @ward1993].

The differences between groups both today and in the past have given rise to much research stratified by racial group. Public health and medical research over the past twenty years has shown that while race is a social construct and thus biological fiction, its social consequences -- namely racism and its potential effects -- have far-reaching influences on health and disease patterns, particularly in countries with high levels of race-based inequality [@meloni2022; @travassos2004]. Thus, racism and racist policies also can influence birth outcomes. Contemporary studies in Brazil have shown that non-White mothers have higher odds of giving birth to LBW infants [@nyarko2013; @wehby2015a; @vilanova2019]. Non-White mothers also had infants with lower mean birth weights than their White counterparts [@leal2006]. These trends also hold true in the United States, the other major slave society in the Americas, where today Black newborns are at higher risk for LBW than White newborns, with trends worsening over time [@pollock2021]. Racism is a possible upstream contributor, as studies demonstrate that U.S.-born Black women have higher risk for both LBW and PTB than Black immigrant women [@braveman2021a; @david1997]. Differences in birth weight and childhood growth patterns among differently racialized groups in South Africa since the end of Apartheid also show a gap between White and Black and "colored" children that starts at birth and increases over time even as all children have gotten larger [@cameron2003].

@@

Brazil was also the last country to abolish slavery (1888) in the Western Hemisphere. In the nineteenth century, population-level health was poor for all Brazilians, but particularly so for enslaved individuals, who experienced the epidemic and endemic diseases such as cholera, tuberculosis, and yellow fever that affected the entire population [@karasch1987]. Although caloric intake for the enslaved population was probably sufficient in terms of quantity, excluding pregnant and nursing women, scholars have shown how it lacked key nutrients, and thus the enslaved suffered from nutritional deficiencies, including thiamine deficiency, which was associated with high infant mortality rates [@karasch1987; @kiple1989]. Today, thiamine deficiency is correlated with lower birth weight and adverse perinatal outcomes including neonatal death [@gr2024; @kareem2023].

@@

By the 1920s, in Rio de Janeiro public maternity hospitals, established in the previous three decades, had grown in number and size as obstetricians and public health officials worked to provide prenatal care and hospitalize labor and delivery [@barreto2016; @roth2020]. Maternidade Laranjeiras, which opened in 1904 and was the teaching hospital for the obstetrics and gynecology program at Rio de Janeiro’s medical school, provided gynecological and obstetric care free of charge; its clientele came from the city’s poor and working classes [@barreto2016; @roth2020]. Most patients were women of color defined as either Black or mixed race, and most White patients were immigrants, the majority from Europe. Because most clinical notes were short – only those that described surgical births such as cesarean sections included details – they do not show if racist thought affected how physicians, all male and mostly White, practiced medicine [@roth2021]. Although the notes show no direct evidence of racist medical practice, the hospital space was one of gender and racial hierarchy – not separate from Brazilian society but its microcosm.

This study builds on an important body of work in economic history that uses hospital records to explore historical trends in birth weights in major U.S. and Western European cities [@costa1998; @costa2004; @goldin1989; @steckel1986; @steckel1998; @ward1993]. It also dialogues with studies in the human sciences that hypothesize an intergenerational effect of the poor health conditions, including poor nutrition and excessive workloads, experienced by enslaved individuals on their descendants across multiple generations [@jasienska2009; @leimert2020; @martins2024].

@@

This study analyzes infant birth weight as a proxy for general fetal health in relation to maternal skin color (Black, mixed-race, White), controlling for maternal age, parity (nulliparous, multiparous), and nationality (European, Latin American, Middle Eastern, Brazilian) in the Maternidade Laranjeiras. It aims to provide some basic quantitative associations of the health status of the descendants of enslaved individuals in post-abolition Brazil to demonstrate the failures of the state to fully incorporate African-descended individuals into political, economic, and social life and to substantiate the qualitative findings from other researchers on health disparities stemming from slave societies, disparities that existed but did not wholly determine patients' lives.

@@

This analysis draws on a unique data set of 2,845 recorded clinical visits to Maternidade Laranjeiras between 1922 and 1926. I extracted the sample from Brazil’s premier obstetrics and gynecology journal, the *Revista de Gynecologia e d’Obstetricia* (RGO). The journal published obstetricians’ and gynecologists’ clinical observations, analyses of new surgical techniques, and professional proceedings. Between June 1922 and May 1926, RGO also published the monthly clinical reports of all women treated at Laranjeiras. I was unable to locate vol. 18, nos. 4, 5, 6, 8 (1924) and vol. 20, no. 4 (1926). In the initial months of publication, the reports included more complete information, including labor and delivery time or detailed descriptions of surgical procedures. Over time, clinical notes became streamlined. When physicians intervened in birth, the notes included the type of intervention, the indication, and the attending obstetrician [@roth2020; @roth2020a].

@@

*Birth weight*. The outcome was birth weight measured as a continuous variable in grams for live births only. In a secondary analysis available in the supplement, I also evaluated birth weight as a binary variable (either normal birth weight, NBW $>2500$ g or low birth weight, LBW $\leq2500$ g), according to current WHO classifications) [@worldhealthorganization2022]. Wilcox (@wilcox2001) has questioned the utility of the LBW category in epidemiological studies given its arbitrary nature. We can question the use of this cutoff in historical studies as well, as it was not an established category until the mid-twentieth century. A Finnish pediatrician first proposed the 2500 g cutoff in 1919 without any specific justification [@hughes2017]. This was a period in which it was difficult for physicians to accurately calculate gestational age, and thus many physicians used birth weight as a proxy for prematurity. The WHO adopted this cutoff in 1948, and it has remained the standard ever since even as epidemiologists began to differentiate between small-for-gestational age (SGA, now FGR) and PTB in the 1970s [@hughes2017; @wilcox2001]. Physicians in early twentieth-century Rio de Janeiro defined low birth weight as BLANK [@conrado1924a].

@@

*Maternal nationality*. The original hospital records recorded maternal nationality by country, which are available in the supplemental file. For analysis here, I created categories based on individual country (`nationality`) and a dummy variable based on region (`modifiednationality`). I recategorized the latter as follows: `Brazilian`; `Latin American` (Argentine, Paraguayan, Uruguayan); `European` (Austrian, French, German, Italian, Polish, Portuguese, Romanian, Russian, Spanish, and Swiss); and `Middle Eastern` (Syrian). For a sub-sample analysis to test collinearity between nationality and skin color, I further created a dummy category between `Brazilian` and `non-Brazilian`.

@@

*Infant length*. Infant length was recorded in centimeters in the original records. A more complete analysis would further explore interactions between birth weight and birth length, which can be a proxy for the nutritional and health status of the mother later in pregnancy [@santana2021]. The third trimester is when most fetal weight gain occurs, and maternal malnutrition during this period can have a dramatic effect on infant birth weight [@jasienska2009; @ward1993]. But other health problems related to maternal undernutrition during pregnancy depend on gestational period. Research on the Dutch famine during World War II has shown that the adult health problems of infants born to mothers who were pregnant during the famine depended on the gestational period during which maternal undernutrition occurred [@roseboom2006; @painter2005].

@@

Scholars have shown that immigrants to Brazil during this historical period were healthier and better off socio-economically than their Brazilian counterparts of all skin colors [@franken2019; @ball2020]. Given the possibility for collinearity between maternal nationality and maternal skin color, I ran a sub-sample analysis for Models 1 and 2 (modified ancestry) in two steps. First, I created a new dichotomous variable, `Brazilian` (yes/no) and ran a simple linear analysis with birth weight as the outcome and maternal nationality as the independent variable. Non-Brazilian women birthed infants that weighed on average 100 g more than Brazilian women (the reference group) (p-value <0.001). Given that all women of color in the data set were Brazilian, collinearity exists. Second, I dropped all non-Brazilians from the study, and ran the original simple and multiple regressions from Models 1 and 2.

@@

The overall rate of LBW for infants born to Afro-descended mothers was 12.4%, compared to 9.7% for Euro-descended mothers. A difference in means test showed borderline significance at the alpha 0.05 level (t-statistic = 1.9119, df. 2008.2, p-value 0.056, [95% CI -0.0007, 0.0532]).

Results for all linear models are displayed in Tables 2 and 3. All models demonstrated an association between maternal skin color and infant birth weight, with Euro-descended women (either as Euro-descended or as White, the reference groups) having infants with higher birth weights than Afro-descended women (either as Afro-descended, Black, or mixed race). When recategorizing maternal skin color into two categories (Afro- and Euro-descended) and running the full model, Afro-descended women had infants who weighed on average 85 grams lighter than their Euro-descended counterparts (Table 2). In Table 3, which shows the models run using the hospital records' original skin color categories, Black mothers gave birth to infants weighing on average 100 grams less than White mothers, while mixed-race mothers gave birth to infants weighing on average 72 grams lighter than White mothers. Infants born to women of African descent were both lighter and more likely to be LBW than infants born to women of only European descent. Research has shown that below 3000 g, risk of death in the first week of life increases by 40% for every 100 g decrease in birth weight [@stein1984; @wilcox1983]. More recent studies relying on IV have found that a 10% increase in birth weight decreases infant mortality by 13.3 deaths per 1000 live births [@maruyama2020]. Applied to these data, 85 g difference in birth weight between Afro- and Euro-descended infants (sub-model 2) would have increased infant mortality among Afro-descended children by 3.7 deaths per 1000 births.

In all models (Tables 2 and 3), women who were nulliparous (delivering their first child) had lighter infants weighing than multiparous women, an association recognized by physicians at the time and found in both other historical studies and today [@conrado1924a; @costa1998; @souza2020]. Age and nationality were not significant.

@@

This study found that women of African descent in early twentieth-century Brazil both had infants with lower birth weights and higher rates of LBW than their White counterparts, perhaps indicative of higher rates of PTB. Regardless of a difference between infants born to mothers of different skin colors, for all LBW infants in the past, the probability of surviving the neonatal period was lower than today given higher neonatal and infant mortality rates and less effective medical interventions [@costa1998; @roth2020]. This fact is born out by more recent time-series analysis of neonatal survival [@maruyama2020].

Table 4 provides the mean birth weights for the Laranjeiras sample, stratified by sex and maternal skin color in comparison with both current-day birth weights from various regions in Brazil and historical studies on birth weight across the Americas and in New Zealand. Black infants born at Laranjeiras had the lowest mean birth weight (3037 g) when analyzed in conjunction with other historical samples except for estimated enslaved birth weights in the antebellum US South [@steckel1986], which scholars have since called into question for being too low [@costa2004]. Black infants born at Laranjeiras were almost 150 g lighter than Black infants born at the Johns Hopkins hospital during an equivalent historical period. They were less than 100 g lighter than Black infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. White infants born at Laranjeiras had a mean birth weight of 3133 g, slightly more than 50 g lighter than White infants born to mothers with incomplete K-12 schooling in early twenty-first century Rio de Janeiro. They were nearly 300 g lighter than White infants born at the Johns Hopkins hospital during an equivalent historical period (3423 g). As other scholars have shown, mean birth weights in the past were not drastically lower than today, and Laranjeiras fits this pattern, with birth weights relatively close to those of contemporary Brazilian infants even if lower than infants born in the United States or New Zealand during the same historical period [@costa1998; @costa2004; @goldin1989].

When compared both historically and contemporaneously, we can see that women at Laranjeiras did not fair as well as their counterparts in other major cities across the globe or with their counterparts today. Although the mean birth weight at Laranjeiras was within today's normal range, the differences between racialized groups demonstrate differential health outcomes that continue to persist. Across populations, relatively high mean birth weights in the nineteenth and early twentieth centuries are probably an artifact of high stillbirth and neonatality rates [@costa1998; @costa2004]. Today, PTB and lower birth weight infants that probably would have died in utero 100 years ago are now surviving, bringing the mean birth weight down. As our estimates of PTB demonstrate, the difference in mean birth weight is probably a reflection of higher rates of PTB among African-descended women.

What can explain the differences in birth weight historically both between all women at Laranjeiras and those of other historical populations and among different skin-color groups within Laranjeiras? The intrauterine environment is a critical determinant of infant birth weight. Maternal age, parity, birth spacing, and gestational age all condition birth weight. In this study, maternal age was not a significant factor, although multiparous mothers had infants with higher birth weights, as is common today. The data did not provide information on birth spacing or gestational age, the latter which I discuss below in the limitations.

What about external factors related to poverty? Infectious disease, alcohol or drug use, smoking, heavy physical work, and inadequate prenatal care also influence birth weight [@stein1984]. Maternal nutrition can as well, although the relationship is nonlinear in that the pregnant body protects the fetus at the expense of the mother, and a threshold of malnutrition must be met before there is an adverse effect on fetal growth [@costa1998; @steckel1986; @stein1984]. Yet nutritional deficiencies that were found among African-descended populations in nineteenth-century Rio de Janeiro due to slavery are associated with lower birth weights and infant mortality [@karasch1987; @kiple1989; @gr2024; @kareem2023]. Although working women of all skin colors had more control over their diets than did the enslaved, poor and working-class Cariocas (as residents of Rio de Janeiro are called) continued to rely heavily on manioc and dried meat, both deficient in thiamine. [CITE] We know that poorer women in early twentieth-century Rio de Janeiro also performed physically taxing jobs that required them to stand for long periods of time, including as washerwomen, domestic help, and street vendors [@graham1992]. And working-class women of all skin colors often worked through the third trimester as domestic servants [@roth2020]. These same women might have lived in crowded housing arrangements with lack of access to running water [@graham1992; @fischer2008; @roth2020]. Thus, generalized poverty might have influenced women of African descent's nutrition and ability to access prenatal care.

@@

All models based on historical data must contend with limited sample sizes, missing data, and imprecise measures of health and nutrition [@franken2019; @steckel1998; @roberts2014; @ward1993]. Physicians could have introduced systematic measurement error when weighing newborns in Laranjeiras, either through inconsistent practices or inaccurate equipment. The hospital records do not provide information on weighing practices, but published observations from medical students who trained at Laranjeiras show that the norm was to weigh all newborns without clothes on a calibrated scale [@conrado1924a]. Thus, an upward bias to the mean weight through the inclusion of clothing or blankets for all infants seems unlikely as does any systematic bias according to infant skin color in any direction. It seems unlikely that physicians weighed infants of one skin color in a different method than those of another given that this was the premier teaching hospital associated with one of Brazil's only medical schools. Comparing the mean weights to those of other historical data sets (Table 4) shows that mean weights at Laranjeiras were lower than all others except for the estimated weights of slave infants in the antebellum southern United States, showing that if any upward bias existed, weights were even lower than presented here. Additionally, birth weights at Laranjeiras were rounded to the nearest 50 grams. If there was a systematic bias in any direction for any group of infants, it would have had to be at the hospital level since multiple physicians and medical students weighed babies over the five-year period. The theoretical upper bound on bias is 50 grams, and my results show a difference greater than 50 g. Claudia Goldin [@goldin1989] has shown for Philadelphia that heaping did not systematically bias the mean, and I assume the same here.

In this study, a major limitation was the original published clinical notes' exclusion of gestational age. Thus, the birth weight variable captured both smaller babies born at term, whether with or without FGR, and pre-term infants (PTB), or those born before 37 weeks' gestation. We are thus unable to know how many of these infants were low birth weight at term [@falcao2020]. This limits our ability to understand the relationship between maternal variables and infant birth weight as pre-term infants are more likely to have lower birth weights [@wilcox2001; @wilcox2024]. Given the absence of accurate technological techniques for determining gestational age in the past, this limitation is inherent to any historical inquiry into birth weights, although other historical studies have data sets that include gestational age [@costa1998; @costa2004; @goldin1989]. Here, I assume that infants at risk for PTB were also at-risk for lower birth weight.

Today, lower mean birth weights between populations are not necessarily associated with worse health outcomes. In fact, Wilcox [@wilcox2001, p. 1236] has shown that "populations with lighter babies do not necessarily have worse mortality." Yet continued differences between racialized groups hints at the possible role of socio-economic inequity and racism in determining health outcomes. In addition, studies relying on instrumental variable (IV) analysis have shown that even in the recent past in developed countries, the impact of birth weight on adverse perinatal and neonatal outcomes has declined over the past 50 years given advances in medical care [@maruyama2020]. We can imagine that 100 years ago, the association of birth weight, most likely as a proxy for PTB, with adverse neonatal outcomes was even greater. Finally, birth weights studied historically are as important for understanding their mother's past lives as for predicting perinatal or neonatal outcomes [@ward1993].

@@
